Dynamic Stochastic Approximation Monte Carlo

### 2018-06-25 papers/histogram/sad.tex
changed gamma to use ratcheting energies.
edits to paper (should we merge methods with history bits of introduction?)

\section{Methods}

In this work, We compare a variety of broad histogram methods.  We
outline the general workings of each algorithm that we developed in
detail while summarizing algorithms that originated elsewhere.  The
following methods are introduced and applied to the square-well fluid:
Transition Matrix Monte-Carlo (TMMC), Wang-Landau (WL), Wang-Landau
Transition Matrix Monte-Carlo (WL-TMMC), Stochastic Approximation
Monte-Carlo (SAMC), and Dynamic Stochastic Approximation
(SAD).

@@

\subsection{Wang-Landau Algorithm}

The Wang-Landau Algorithm developed by Wang and Landau proposes a
random walk in energy space in order to estimate the density of
states~\cite{wang2001efficient,wang2001determining, landau2014guide}.
Wang-Landau's major tenant is that when counting the histogram
$H(\epsilon)$, the energy occurrences should form a flat distribution.
The criteria for flat sampling is given by
\begin{equation}
	\frac{\min_{\epsilon} H(\epsilon)}
	{\big\langle H(\epsilon)\big\rangle }
	> \eta
\end{equation}
where $\eta$ is usually set to $0.80$ and determines how many times
each energy is sampled relative to the mean number of visits.
Difficulties can occur with this flatness criteria due to the fact that
some energies on a given energy range might never be
sampled~\cite{haber2014transition}. The algorithm is briefly outlined
below where $w$ represents the weights:
\begin{equation}
	\ln{D_{i+1}(\epsilon,N)}=\ln{D_{i}(\epsilon,N)}
	+\gamma
\end{equation}
\begin{equation}
	\gamma_{i+1}=\frac{u}{2} \gamma_{i}
\end{equation}
for $u$ typically greater than 1 and $\gamma$ is the modification
factor for the density of states.  A new configuration for the density
of states with a probability given by the following is chosen:
\begin{equation}
	\mathcal{P}(\epsilon_\text{old} \rightarrow \epsilon_\text{new})
	= \min[1,e^{\ln{D(\epsilon_\text{old})}-\ln{D(\epsilon_\text{new})}}]
\end{equation}
The entire process is repeated until a desired cutoff is reached.

@@

\subsection{SAMC Algorithm}
SAMC can be applied as a dynamic importance sampling algorithm that
uses an update factor to explore energy
space~\cite{liang2007stochastic, werlich2015stochastic,
schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$ and a scale factor $\gamma_0$.
\begin{align}
\gamma_{t}^{\text{SA}} =\gamma_0 \frac{t_0}{\max(t_0,t)}
\end{align}
SAMC offers several advantages over other broad histogram methods
including proven convergence. It has been shown that if
$F_{t}^{\text{SA}}$ satisfies two conditions the estimator is proven to
converge to the real DOS~\cite{liang2006theory, liang2007stochastic}.
\begin{align}
\sum_{t=1}^\infty \gamma_{t}^{\text{SA}} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty (\gamma_{t}^{\text{SA}})^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$.  Unlike WL and WL-TMMC, the energy range need
not be known a priori.  The time to converge depends only on the choice of
parameters $t_0$ and $\gamma_0$.
Unfortunately, only by performing many computational runs might the user
determine what SAMC arbitrary parameters to choose.  This has lead us
to develop a systematic way (which we will discuss in the next section)
for the simulation to determine what $t_0$ should be.

\section{SAD Algorithm}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) method is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Broad-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$ which is an immediate disadvantage of the method;
however, this allows the user to define an energy simulation
range between $T_\text{min}$ and $\infty$. Setting a physical parameter
such as a minimum temperature $T_\text{min}$ is intuitive vs. a user
needing to know a prior either an energy range or some unphysical
parameter $t_0$.

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SA}}$ can
be thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.  We argue that we can write
this entropy `term' geometrically as an area bounded by the $\ln{D(E)}$
i.e. the entropy. This gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas
bounded by the $\ln{D(E)}$ and the minimum temperature line. The update
factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. In contrast to SAD, Fig.~\ref{fig:gamma-vs-t}. shows the
monotonic decrease of the update factor for SAMC.  The update factor
for WL decreases like a step function while the update factor for SAD
dynamically fluctuates around a value less than 1 for early MC moves
(while finding new energy states).  Ultimately once all energy states
have been found, the update factor for SAD decreases in the same
monotonic fashion as SAMC.

>>>>

\davidsays{Fix this, it was moved when I reorganized the methods
  section} In this work, We compare a variety of broad histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that originated
elsewhere.  The following methods are introduced and applied to the
square-well fluid: Transition Matrix Monte-Carlo (TMMC), Wang-Landau
(WL), Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC), Stochastic
Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation
(SAD).

\section{Broad histogram methods}

The goal of broad histogram methods (also called \emph{flat histogram}
or \emph{multicanonical} methods) is to simulate each energy with
similar accuracy, so as to accurately determine the density of states
over a broad range of energies.  Once the density of states is known,
a number of thermodynamic quantities---such as heat capacity or
internal energy---can be easily computed for any temperature.
Properties that require more information---such as a spatial
correlation function or a response function---can still be computed
for any temperature, provided statistics are collected for each
individual energy, which can then be reweighted for any temperature.

Each broad histogram Monte Carlo method uses a given set of ``moves''
which change a system and should satisfy detailed balance.  Each
algorithm differs in how it determines the probability of accepting a
move, and in what additional statistics (if any) must be collected in
order decide on that probability.  The methods we will compare fall in
two categories:  weight-based methods, which only require a set of
weights $w(E)$ be collected, and transition-matrix methods that
require that statistics be collected for any transitions between pairs
of energies.

\davidsays{We should discuss continuous energies and binning somewhere
  here...}

\section{Weight-based methods}

We will begin by introducing two related weight-based methods which
rely on a weight function $w(E)$, noting that the method proposed in
this paper (discussed in Section~\ref{sec:sad}) also falls in this
category.  In these algorithms, the probability of accepting a move is
given by
\begin{equation}
	\mathcal{P}(\epsilon_\text{old} \rightarrow \epsilon_\text{new})
	= \min\left[1,\frac{w(\epsilon_\text{old})}{w(\epsilon_\text{new})}\right]
\end{equation}
This biases the simulation in favor of energies with low weights.  A
set of weights that are proportional to the density of states of the
system $D(\epsilon)$ will result in an entirely flat histogram.  Thus
the goal of each of these methods is to achieve a flat histogram.  For
numerical reasons, it is common to store and work with the natural
logarithm of the weights, since the weights may vary over more than a
few hundred orders of magnitude.  Note that the logarithm of the
density of states is the entropy at that energy, so we can think of
the logarithm of the weights as an approximation of the entropy
\davidsays{make this less colloquial}.

Each weighted density approach use a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(\epsilon)}=\ln{w_{t}(\epsilon)}
	+\gamma
\end{equation}
where $t$ is number of the current move, and $\gamma$ is an update
factor.  This update causes the random walk to avoid energies that
have been frequently sampled, leading to a rapid exploration of energy
space.  This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreased as we decrease $\gamma$.  The various
weight-based methods discussed here differ in how the schedule the
decreasing of $\gamma$.

\subsection{Wang-Landau}

Wang-Landau's approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma=1$, and then decreases $\gamma$
in discrete stages.  We track the number of steps at each energy during
each stage in a histogram.  When that histogram is sufficiently flat,
$\gamma$ is decreased by a specified factor, which is usually
$\frac12$.  The flatness is defined by the ratio between the minimum
value of the histogram and its average value.  When this flatness
reaches a specified threshold (typically 0.8), the $\gamma$ value is
decreased.  This approach requires that the energy range of interest
be known in advance, and difficulties can occur with this flatness
criteria due to the fact that some energies in this energy range might
never be sampled~\cite{haber2014transition}.  The entire process is
repeated until $\gamma$ reaches a desired cutoff is reached.

The Wang-Landau approach thus has three parameters that need be
specified: the factor by which to decrease $\gamma$ when flatness is
acheived, the flatness criterion, and the cutoff that determines when
the computation is complete.  In addition, an energy range (or more
specifically, a set of energies) must be supplied, so that the
flatness criterion can be defined.

\davidsays{We need to look up these references.}  While this approach
is very efficient and has been widely uses, it suffers a few
shortcomings.  Firstly, the set of energies must be specified, which
may require multiple simulations.  Secondly, while Wang-Landau
converges quickly, it does not in general converge to the true density
of states.  It can and does decrease $\gamma$ so quickly that it will
never (for any cutoff value) decrease the error in the density of
states beyond a given value.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$ and a scale factor
$\gamma_0$. \davidsays{Did it originally have a scale factor?}
\begin{align}
\gamma_{t}^{\text{SA}} =\gamma_0 \frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameters $t_0$ and $\gamma_0$.  Unfortunately, there is no
prescription for finding acceptable values for these parameters, and
while the algorithm formally converges, for a poor choice of $t_0$
that convergence may be far too slow to be practical.  This work
proposes a means for dynamically setting the update factor $\gamma_t$
that should robustly converge on a variety of systems.

\section{Transition-matrix methods}

@@

@@

\section{SAD Algorithm}\label{sec:sad}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) method is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Broad-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$ which is an immediate disadvantage of the method;
however, this allows the user to define an energy simulation
range between $T_\text{min}$ and $\infty$. Setting a physical parameter
such as a minimum temperature $T_\text{min}$ is intuitive vs. a user
needing to know a prior either an energy range or some unphysical
parameter $t_0$.

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SA}}$ can
be thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.  We argue that we can write
this entropy `term' geometrically as an area bounded by the $\ln{D(E)}$
i.e. the entropy. This gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{\epsilon_{H}-\epsilon_{L}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas
bounded by the $\ln{D(E)}$ and the minimum temperature line. The update
factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. In contrast to SAD, Fig.~\ref{fig:gamma-vs-t}. shows the
monotonic decrease of the update factor for SAMC.  The update factor
for WL decreases like a step function while the update factor for SAD
dynamically fluctuates around a value less than 1 for early MC moves
(while finding new energy states).  Ultimately once all energy states
have been found, the update factor for SAD decreases in the same
monotonic fashion as SAMC.
